Estimation below the detection limit with application to cryo-EM

### 2018-08-13 TEX/inverse_problem_draft.tex
inverse problem draft

\title{Estimation below the detection limit with application to cryo-em}

@@

We consider the problem of estimating a signal from its multiple occurrences in a measurement. In the high SNR regime, the problem is easy. In the low SNR, it becomes interesting and that is the focus of this paper. 
The problem can be posed as blind deconvolution between the signal and a sparse binary signal.
We can also consider the extension for multiple signals.

The main motivation for this paper arises from single particle reconstruction using cryo-electron microscopy (cryo-EM). In cryo-EM, multiple samples .....

in low SNR detection is impossible. This led to the belief in the cryo-EM community is reconstruction of small molecules is impossible. We claim that this argument is wrong. While this paper focused on the mathematical toy model, Section TKTK discusses the application to cryo-EM. A companion paper focused on this application and provides all the details~\cite{bendory2018toward}.

@@

Let $x_1,\ldots,x_K\in\RL$ be the sought signals and let $y\in\RN$ be the observed micrograph (notice that it is equivalent to think of the data as being one long micrograph or multiple shorter micrographs concatenated into one.)
The forward model (or ``image'' formation model) is as follows. For each target signal $x_k$, an unknown binary signal $s_k \in \{0, 1\}^{N-L+1}$ indicates (with 1's) the starting positions of all occurrences of $x_k$ in $y$, so that, with additive white Gaussian noise:
\begin{align}
y & = \sum_{k=1}^K x_k \ast s_k + \varepsilon, & \varepsilon & \sim \mathcal{N}(0,\sigma^2 I_N),
\label{eq:model}
\end{align}
where $\ast$ denotes linear convolution. 
We consider two models for the generative process of the sparse signal:

\paragraph{The well separated support model.}
The binary signals obey the following property:
\begin{align}
\textrm{If } s_k[i] = 1 \textrm{ and } s_{k'}[j] = 1 \textrm{ but } (k, i) \neq (k', j), \textrm{ then } |i - j| \geq 2L-1.
\label{eq:spacing}
\end{align}
In words: the starting positions of any two occurrences (be it of the same signal or of two different signals) must be separated by at least $2L-1$ positions, so that their end points are necessarily separated by at least $L-1$ signal-free entries in the micrograph.

\paragraph{Poisson process model.} In this model, points are chosen in $\{1,\dots,n\}$ according to a Poisson process with parameter $\gamma n$. For each point $i$ that is chosen from $1$ to $n$, a random vector $X$ from the distribution is then placed in the large vector, with element $0$ at location $i$, with overlapping vectors being added together.

If $M_i$ denotes the number of hits at location $i$, $1 \le i \le n$, then by definition of the Poission process $M_i$'s are i.i.d.\
 and $M_i \sim \Poisson(\gamma)$. Conditional on the value of $M = (M_1,\dots,M_n)$, if we let $X_1^{i},\dots,X_{M_i}^i$ denote the random vectors with position 0 located at $i$, then $X_{k_1}^{i}$ and $X_{k_2}^{i}$ are independent for $k_1 \ne k_2$.

From $y$, we aim to recover $x_1, \ldots x_K$, and possibly also the number of occurrences of each. In contrast, particle picking is the task of estimating the binary signal $s = s_1 + \cdots + s_K$, while clustering is the task of separating $s$ into its components $s_1, \ldots, s_K$. Neither of the latter can be performed reliably if $\sigma$ is large (that is, at low $\SNR$.)

%We denote the set of these nonzero values by 
% $\mathcal{S}_i$ and its cardinality by $\vert \mathcal{S}_i\vert = M_i$. 
%By assuming that all $\mathcal{S}_i$'s are disjoint,  we let $s = \sum_{i=1}^Ks_i$, $\mathcal{S} = \bigcup_{i=1}^{K} \mathcal{S}_i$ and  $\vert \mathcal{S}\vert :=M =  \sum_{i=1}^{K}M_i$.  %Neither the $M_i$'s nor $M$ are assumed to be known.

Considering~\eqref{eq:model}, one can think of this inverse problem as a mixture of blind deconvolution problems between binary signals and the target signals. %Related literature is briefly surveyed in Appendix~\ref{sec:related_literature}.

@@

For $K=1$, our problem can be interpreted as a special case of the system identification problem. Similarly to~\eqref{eq:model}, the system identification forward model takes the form
%
\begin{math}
%
y = x\ast w + \varepsilon,  
%
\end{math} 
%
where $x$ is the unknown signal (the ``system''), $w$ is an unknown, random, input sequence, and $\varepsilon$ is an additive noise.   
%The problem has also been studied in the case of a known input $w$~\cite{pillonetto2010new,dinuzzo2015kernels,bottegal2016robust}. 
The goal of this problem is to estimate $x$, usually referred to as ``identifying the system.'' The question of identifiability of $x$ under this observation model is addressed for certain Gaussian and non-Gaussian $w$ in~\cite{benveniste1980robust,kormylo1983identifiability}. In the special case where $w\in\{0,1\}^N$, satisfying the spacing requirement~\eqref{eq:spacing}, we obtain our model in the  case of a single signal ($K = 1$). The same observation model is used for blind deconvolution, a longstanding problem arising in a variety of engineering and scientific applications such as astronomy, communication, image deblurring, system identification and optics; see~\cite{jefferies1993restoration,shalvi1990new,ayers1988iterative,abed1997blind}, just to name a few. 
%\TODO{ref this by Giannakis: \cite{giannakis1989identification}}

@@

In special cases, including the case where $w$ is binary, expectation maximization (EM) has been used~\cite{cappe1999simulation}. The EM method for discrete $w$ is based upon a certain ``forward-backward'' procedure used in hidden Markov models~\cite{rabiner1989tutorial}. However, the complexity of this procedure is nonlinear in $N$, and therefore its usage is limited for big data sets.

Indeed, on each iteration of EM, a probability must be assigned to any feasible combination of positions for the current signal estimate in $M$ locations on the grid $\{1,\ldots,N\}$.
In total, even when excluding forbidden combinations due to the spacing constraint, there are $O(N^M)$ such combinations, and the problem becomes computationally intractable when $M$ grows with $N$ and $N$ is large.

@@

\TODO{The following was copied from the appendix of the old draft}

Many automatic and semi-automatic methods for particle picking have been proposed, based on edge detection, template matching and deep learning; see for instance~\cite{harauz1989automatic,ogura2004automatic,zhu2016deep,frank1983automatic,scheres2015semi,heimowitz2018apple}. 
However, most of these procedures are prone to \emph{model bias}. For instance, in the popular framework of RELION~\cite{scheres2015semi}, the user manually marks hundreds of spots on the micrograph, believed to contain projections. 
Therefore, the algorithm's performance depends on the prior assumptions of the users about the particle's structure; the same holds true for deep learning based approaches which require constructing labeled sets of data.
Other methods use disks or differences of Gaussians as templates~\cite{langlois2014automated,voss2009dog}.
Nowadays, it also still popular to pick particles manually. This method, while it exploits the researcher's experience, is both tedious and subject to model bias.

@@

The spacing constraint~\eqref{eq:spacing} gives rise to more powerful observations. Consider the second-order autocorrelation in particular: $a_y^2[\ell]$ computes the correlation between $y$ and a copy of $y$ shifted by $\ell$ entries. Considering $\ell$ only in the range $0, \ldots, L-1$, one can see that any given occurrence of $x$ in $y$ is only ever correlated with itself (with the same shift $\ell$), and never with another occurrence. As a result,
\begin{align*}
\lim_{N\to\infty} a_y^2[\ell] & = \gamma a_{x}^2[\ell] + \sigma^2\delta[\ell]
\end{align*}
for $\ell = 0, \ldots, L-1$, where $\delta$ denotes the Kronecker delta function. The last part captures the autocorrelation of the noise. Notice that, even if $\sigma$ is unknown, entries $\ell = 1, \ldots, L-1$ still provide useful information about $a_x^2$.
Along the same lines, one can establish a relation for third-order autocorrelations:
\begin{align}
\lim_{N\to\infty} a_y^3[\ell_1,\ell_2] & = \gamma a_{x}^3[\ell_1,\ell_2] + \sigma^2\gamma a_{x}^1 \cdot \big(\delta[\ell_1,0]+\delta[0,\ell_2]+\delta[\ell_1,\ell_2]\big),
\label{eq:data_ac_k1}
\end{align}
for $\ell_1,\ell_2 = 0, \ldots, L-1$. Here too, few entries are affected by $\sigma$ in the limit.
Detailed derivations for identities in this and the next part are given in Appendix~\ref{sec:autocorrelation_computation}.

@@

Returning to the general case $K \geq 1$, let $M_1, \ldots, M_K$ denote the number of occurrences of signals $x_1, \ldots, x_K$ respectively, and define
\begin{align}
\gamma_k & = \frac{M_k L}{N}, & \gamma & = \sum_{k=1}^K\gamma_k.
\end{align}
As above, we consider the asymptotic regime where $M_1,\ldots,M_K,N\to\infty$ while preserving the ratios $\gamma_k$ constant.
% 
Still under the spacing constraint~\eqref{eq:spacing}, similarly to the developments above, one can estimate a mixture of the autocorrelations of the $K$ target signals from the autocorrelations of the micrograph:
\begin{align}
\lim_{N\to\infty} a_y^1 & = \sum_{k=1}^K\gamma_k a_{x_k}^1, \nonumber\\
\lim_{N\to\infty} a_y^2[\ell] & = \sum_{k=1}^K\gamma_k a_{x_k}^2[\ell] +\sigma^2\delta[\ell],  \label{eq:data_ac}\\
\lim_{N\to\infty} a_y^3[\ell_1,\ell_2] & = \sum_{k=1}^K\gamma_k a_{x_k}^3[\ell_1,\ell_2] + \sigma^2\left(\sum_{k=1}^K\gamma_k a_{x_k}^1\right)(\delta[\ell_1,0]+\delta[0,\ell_2]+\delta[\ell_1,\ell_2]), \nonumber
\end{align}
where $\ell, \ell_1, \ell_2 = 0, \ldots, L-1$. The left hand side is straightforward to estimate from data: it provides a succinct summary of it. The right hand side involves polynomial functions of unknowns $\gamma_1, \ldots, \gamma_K, x_1, \ldots, x_K$, and possibly $\sigma^2$. The task is to solve these polynomial equations in a robust way.

@@

\begin{proposition} \label{prop:poisson_model}
Under the Poisson process model, the first autocorrelation  is $\gamma \LL_1$, the second moment vector is $(\gamma \LL_1)^2 + \gamma \LL_2(\Delta)$, and the third moment matrix is $(\gamma \LL_1)^3 + \gamma \LL_1  \cdot ( \gamma\LL_2(\Delta_1) + \gamma\LL_2(\Delta_2) + \gamma\LL_2(\Delta_2-\Delta_1)) + \gamma \LL_3(\Delta_1,\Delta_2) $. In particular, from the first three moments of the Poisson process model, once can recover the first three moments from the strongly-separated model, with the Poisson rate $\gamma$ playing the role of the ``occupancy factor''. So if recovery is possible for the strongly-separated model, it is also possible for the Poisson process model.
\end{proposition}

@@

Furthermore,
as proven in the following simple proposition.
%
\begin{proposition} \label{prop:uniqueness}
%
Let $z\in\RL$ and suppose that $z[0]$ and $z[L-1]$ are nonzero. Then $z$ is determined uniquely from  $a_z^2$ and $a_z^3$. More precisely, suppose we can measure $\tilde{a}_z^3[k,L-1] = a_z^3[k,L-1]+\upsilon$ and that $\vert z[0]z[L-1]\vert \geq \delta>0$.
Then,  $\hat{z}[k] =\frac{\tilde{a}_z^3[k,L-1]}{a_z^2[L-1]} $ satisfies $\vert \hat{z}[k] - z[k]\vert\leq \frac{\vert \upsilon\vert }{\delta}$.

@@

\begin{proof}
%
By assumption $a_z^2[L-1] = z[0]z[L-1]\neq 0$.
Then, the uniqueness results, for all $k=0,\ldots L-1$,  follows from:
%
\begin{equation*}
%
a_z^3[k,L-1] = z[0]z[k]z[L-1].
%
\end{equation*}
%
In addition, 
%
\begin{equation*}
%
\hat{z}[k] = \frac{\tilde{a}_z^3[k,L-1]}{a_z^2[L-1]} = z[k]+\frac{\upsilon}{a_z^2[L-1]} \quad \Rightarrow \quad \vert \hat{z}[k] - {z}[k]\vert \leq \frac{\vert\upsilon\vert}{\delta}.
%
\end{equation*} 
%
\end{proof}

@@

%\TODO{TB: I would place the equation counting argument in the theory section in the paragraph of open questions (I marked th place)}
%
Given the observation $y$, we proceed to compute the autocorrelations. The first-order autocorrelation is straightforward. For second-order autocorrelations, notice from equation~\eqref{eq:data_ac} that $a_y^2[\ell]$ suffers no noise-induced bias for $\ell$ in $1$ to $L-1$. Thus, we omit $\ell = 0$, which has the practical effect that we need not know $\sigma$ to make sense of the computed quantities. Likewise, for third-order autocorrelations, $a_y^3[\ell_1, \ell_2]$ for $0 \leq \ell_1, \ell_2 \leq L-1$ such that $\ell_2 \leq \ell_1$ includes all relevant entries for our purpose (this accounts for symmetries), and we further exclude any such that $\ell_1, \ell_2$ or $\ell_1 - \ell_2$ are zero to avoid the need to estimate $\sigma$---there are $\frac{(L-1)(L-2)}{2}$ remaining entries. We have
\begin{align*}
1 + (L-1) + \frac{(L-1)(L-2)}{2} = \frac{1}{2} L (L-1) + 1
\end{align*}
coefficients in total. Since we aim to estimate $KL$ parameters (for the $K$ signals of length $L$) plus $K$ parameters (for the densities $\gamma_k$), an absolute upper bound on $K$ (simply to ensure we have at least as many equations as we have unknowns) is
\begin{align*}
K(L+1) \leq \frac{1}{2} L (L-1) + 1.
\end{align*}
Thus, $(L-1)/2$ (up to a small approximation) is an absolute upper limit on $K$. \TODO{We may want to cite MRA literature here.}

In practice, the autocorrelations are computed on disjoint segments of $y$ of length $100\cdot10^6$ and added up, without correction for the junction points. Segments are handled sequentially on a GPU, as GPUs are particularly well suited to execute simple instructions across large vectors of data. If multiple GPUs are available, segments can of course be handled in parallel.

@@

\TODO{Should we discuss CTF? Where, and to what extent?}

@@

\paragraph{Open theoretical questions.}

@@

We will also denote by $\L_l$ the autocorrelations of $X$:
%
\begin{align}
%
\LL_1 = \sum_{i=0}^{L-1} \M_1[i],
%
\end{align}
%
\begin{align}
%
\LL_2(\Delta) = \sum_{i=0}^{L-1} \M_2[i,i+\Delta],
%
\end{align}
%
and
%
\begin{align}
%
\LL_3(\Delta_1,\Delta_2) = \sum_{i=0}^{L-1} \M_3[i,i+\Delta_1,i+\Delta_2].
%
\end{align}

Note that in the strongly-separated model, the first three observed moments are, respectively, $\LL_1$, $\LL_2(\Delta)$, and $\LL_3(\Delta_1,\Delta_2)$.

@@

In this notation, we will show that the first moment of the data is $\gamma \LL_1$, the second moment vector is $(\gamma \LL_1)^2 + \gamma \LL_2(\Delta)$, and the third moment matrix is $(\gamma \LL_1)^3 + \gamma \LL_1  \cdot ( \gamma\LL_2(\Delta_1) + \gamma\LL_2(\Delta_2) + \gamma\LL_2(\Delta_2-\Delta_1)) + \gamma \LL_3(\Delta_1,\Delta_2) $. In particular, from the first three moments of the Poisson process model, once can recover the first three moments from the strongly-separated model, with the Poisson rate $\gamma$ playing the role of the ``occupancy factor''. So if recovery is possible for the strongly-separated model, it is also possible for the Poisson process model.

@@

\section{The first moment of $Y$}

@@

\section{The second moment of $Y$}

Again, we will condition on $M$ first, and then take the expectation over $M$. Fix $i_1 \ne i_2$, and let $\Delta = i_2 - i_1$. Then:
%
\begin{align}
%
Y_{i_1} Y_{i_2} 
&= \sum_{j_1=0}^{L-1} \sum_{j_2=0}^{L-1} 
\sum_{k_1=1}^{M_{i_1-j_1}}\sum_{k_2=1}^{M_{i_2-j_2}}
X_{k_1}^{i_1-j_1}[j_1] X_{k_2}^{i_2 - j_2}[j_2].
%
\end{align}

We break up the double sum over $j_1$ and $j_2$ into two terms: one where $j_2 \ne j_1 + \Delta$, and one where $j_2 = j_1 + \Delta$ or equivalently $i_1-j_1 = i_2-j_2$. In the first case, all the terms are independent, and so the expectation factors. In the second case, when $k_1 \ne k_2$ we have independence, but otherwise not. This gives (all expectations are conditional on $M$):

>>>>

\title{Estimation below the detection limit with application to cryo-EM}

@@

%
%We consider the problem of estimating a signal from its multiple occurrences in a noisy measurement. If the signal-to-noise ratio (SNR) in the measurement is high, then one can try to detect the signal's occurrences and average. However, in a low SNR, detecting the individual occurrences becomes challenging. In this paper, we propose a method to estimate the signal, or even multiple signals, in a low SNR environment using autocorrelation analysis. We show that under two models specified below, the signal can be estimated to any desired accuracy, if it appears sufficiently many times.
%We can also consider the extension for multiple signals.

Let $x\in\RL$ be the target signal and let $y\in\RN$ be the observed data. Let  $s \in \{0, 1\}^{N-L+1}$ be a binary signal indicating (with 1's) the starting positions of all occurrences of $x$ in $y$, so that, with additive white Gaussian noise:
\begin{align}
y & =  x \ast s + \varepsilon, & \varepsilon & \sim \mathcal{N}(0,\sigma^2 I_N),
\label{eq:model}
\end{align}
where $\ast$ denotes linear convolution. 
While both $x$ and $s$ are unknown, the goal is only to estimate $x$ from $y$. The parameters of the signal $s$ (the locations of its nonzero values) are the \emph{nuisance variables} of the problem. As will be shown next, in the low signal-to-noise ratio (SNR) environment, estimating $s$ is impossible, where estimating $x$ is tractable under some conditions on $s$.

The model can be extended to multiple signals. In this setup,  $x_1,\ldots,x_K\in\RL$ are the sought signals and the data is give by
\begin{align}
y & = \sum_{k=1}^K x_k \ast s_k + \varepsilon, & \varepsilon & \sim \mathcal{N}(0,\sigma^2 I_N),
\label{eq:model_heterogeneity}
\end{align}
where  $s_k \in \{0, 1\}^{N-L+1}$. 
From $y$, we aim to recover $x_1, \ldots x_K$, and possibly also the number of occurrences of each.
Considering~\eqref{eq:model}, one can think of this inverse problem as a mixture of \emph{blind deconvolution} problems between binary signals and the target signals.

The generative model of the support signals play a crucial role in the ability to estimate the signals in low SNR. In this paper, we consider two models: one deterministic and one probabilistic.

\paragraph{Deterministic well separated support.}
The binary signals obey the following property:
\begin{align}
\textrm{If } s_k[i] = 1 \textrm{ and } s_{k'}[j] = 1 \textrm{ but } (k, i) \neq (k', j), \textrm{ then } |i - j| \geq 2L-1.
\label{eq:spacing}
\end{align}
In words: the starting positions of any two occurrences (be it of the same signal or of two different signals) must be separated by at least $2L-1$ positions, so that their end points are necessarily separated by at least $L-1$ signal-free entries in the micrograph.

\paragraph{Probabilistic Poisson  model.} In this model, points are chosen in $\{1,\dots,N\}$ according to a Poisson process with parameter $\gamma n$. For each point $i$ that is chosen from $1$ to $n$, a random vector $X$ from the distribution is then placed in the large vector, with element $0$ at location $i$, with overlapping vectors being added together. The single signal case fits a Delta distribution, while the heterogeneous case with $K>1$ implies a sum of Delta distribution on $X$.

If $M_i$ denotes the number of hits at location $i$, $1 \le i \le M$, then by definition of the Poisson process $M_i$'s are i.i.d.\
 and $M_i \sim \Poisson(\gamma)$. Conditional on the value of $M = (M_1,\dots,M_n)$, if we let $X_1^{i},\dots,X_{M_i}^i$ denote the random vectors with position 0 located at $i$, then $X_{k_1}^{i}$ and $X_{k_2}^{i}$ are independent for $k_1 \ne k_2$.

@@

@@

For $K=1$, our problem can be interpreted as a special case of the system identification problem. Similarly to~\eqref{eq:model}, the system identification forward model takes the form
%
\begin{math}
%
y = x\ast w + \varepsilon,  
%
\end{math} 
%
where $x$ is the unknown signal (the ``system''), $w$ is an unknown, random, input sequence, and $\varepsilon$ is an additive noise.   
%The problem has also been studied in the case of a known input $w$~\cite{pillonetto2010new,dinuzzo2015kernels,bottegal2016robust}. 
The goal of this problem is to estimate $x$, usually referred to as ``identifying the system.'' The question of identifiability of $x$ under this observation model is addressed for certain Gaussian and non-Gaussian $w$ in~\cite{benveniste1980robust,kormylo1983identifiability}. In the special case where $w$, satisfying one of the generative models above, we obtain our model in the  case of a single signal ($K = 1$). The same observation model is used for blind deconvolution, a longstanding problem arising in a variety of engineering and scientific applications such as astronomy, communication, image deblurring, system identification and optics; see~\cite{jefferies1993restoration,shalvi1990new,ayers1988iterative,abed1997blind}, just to name a few. 
%\TODO{ref this by Giannakis: \cite{giannakis1989identification}}

@@

In special cases, including the case where $w$ is binary, EM has been used~\cite{cappe1999simulation}. The EM method for discrete $w$ is based upon a certain ``forward-backward'' procedure used in hidden Markov models~\cite{rabiner1989tutorial}. However, the complexity of this procedure is nonlinear in $N$, and therefore its usage is limited for big data sets. 
Indeed, for a general support signal $s$, on each iteration of EM, a probability must be assigned to any feasible combination of positions for the current signal estimate in $M$ locations on the grid $\{1,\ldots,N\}$---$O(N^M)$ such combinations in total. Hence, the problem becomes computationally intractable when $M$ grows with $N$ and $N$ is large. However, exploiting the separation constraint on the support~\eqref{eq:spacing}, we devise in Section TKTK an approximate EM procedure for  our model.

\TODO{Need to refer to this paper https://arxiv.org/abs/1806.00338}

@@

The main motivation for this paper arises from single particle reconstruction using cryo-electron microscopy (cryo-EM).
In a cryo-EM experiment, multiple samples of a molecule are frozen in a thin layer of ice. The microscope is then generates a tomographic image of the ice layer (with the molecules) called micrograph. The goal is to reconstruct the 3-D structure of the molecule from the several micrographs. 
Importantly, the electron dosage transmitted by the microscope must kept low to avoid destruction of the molecules' structure. This, in turn, induces high noise level on the micrograph.

The standard algorithmic pipeline in cryo-EM begins by detecting the projections of the samples in the micrograph. This procedure is called \emph{particle picking}. Then, the 3-D structure of the molecule is constructed from the projections, usually using different variations of expectation-maximization (EM). 
Many automatic and semi-automatic methods for particle picking have been proposed, based on edge detection, template matching and deep learning; see for instance~\cite{harauz1989automatic,ogura2004automatic,zhu2016deep,frank1983automatic,scheres2015semi,heimowitz2018apple}.

If the molecule to be reconstructed is small, then the SNR might be very low. 
As been recognized by cryo-EM scientists, particle picking is impossible in low SNR and thus for small particles. 
This led to the belief in the cryo-EM community is reconstruction of small molecules is impossible. We claim that this argument is wrong. While this paper focused on a mathematical toy model, Section TKTK discusses the application to cryo-EM. A companion paper focused on this application and provides all the details~\cite{bendory2018toward}.

In addition, most of the particle picking procedures are prone to \emph{model bias}. For instance, in the popular framework of RELION~\cite{scheres2015semi}, the user manually marks hundreds of spots on the micrograph, believed to contain projections. 
Therefore, the algorithm's performance depends on the prior assumptions of the users about the particle's structure; the same holds true for deep learning based approaches which require constructing labeled sets of data.
Other methods use disks or differences of Gaussians as templates~\cite{langlois2014automated,voss2009dog}.
Nowadays, it also still popular to pick particles manually. This method, while it exploits the researcher's experience, is both tedious and subject to model bias.

@@

The spacing constraint~\eqref{eq:spacing} gives rise to more powerful observations. Consider the second-order autocorrelation in particular: $a_y^2[\ell]$ computes the correlation between $y$ and a copy of $y$ shifted by $\ell$ entries. Considering $\ell$ only in the range $0, \ldots, L-1$, one can see that any given occurrence of $x$ in $y$ is only ever correlated with itself (with the same shift $\ell$), and never with another occurrence. As a result,
\begin{align*}
\lim_{N\to\infty} a_y^2[\ell] & = \gamma a_{x}^2[\ell] + \sigma^2\ell[\ell]
\end{align*}
for $\ell = 0, \ldots, L-1$, where $\ell$ denotes the Kronecker delta function. The last part captures the autocorrelation of the noise. Notice that, even if $\sigma$ is unknown, entries $\ell = 1, \ldots, L-1$ still provide useful information about $a_x^2$.
Along the same lines, one can establish a relation for third-order autocorrelations:
\begin{align}
\lim_{N\to\infty} a_y^3[\ell_1,\ell_2] & = \gamma a_{x}^3[\ell_1,\ell_2] + \sigma^2\gamma a_{x}^1 \cdot \big(\ell[\ell_1,0]+\ell[0,\ell_2]+\ell[\ell_1,\ell_2]\big),
\label{eq:data_ac_k1}
\end{align}
for $\ell_1,\ell_2 = 0, \ldots, L-1$. Here too, few entries are affected by $\sigma$ in the limit.
Detailed derivations for identities in this and the next part are given in Appendix~\ref{sec:autocorrelation_computation}.

@@

Returning to the general case $K \geq 1$, let $M_1, \ldots, M_K$ denote the number of occurrences of signals $x_1, \ldots, x_K$ respectively, and define
\begin{align}
\gamma_k & = \frac{M_k L}{N}, & \gamma & = \sum_{k=1}^K\gamma_k.
\end{align}
As above, we consider the asymptotic regime where $M_1,\ldots,M_K,N\to\infty$ while preserving the ratios $\gamma_k$ constant.
% 
Still under the spacing constraint~\eqref{eq:spacing}, similarly to the developments above, one can estimate a mixture of the autocorrelations of the $K$ target signals from the autocorrelations of the micrograph:
\begin{align}
\lim_{N\to\infty} a_y^1 & = \sum_{k=1}^K\gamma_k a_{x_k}^1, \nonumber\\
\lim_{N\to\infty} a_y^2[\ell] & = \sum_{k=1}^K\gamma_k a_{x_k}^2[\ell] +\sigma^2\ell[\ell],  \label{eq:data_ac}\\
\lim_{N\to\infty} a_y^3[\ell_1,\ell_2] & = \sum_{k=1}^K\gamma_k a_{x_k}^3[\ell_1,\ell_2] + \sigma^2\left(\sum_{k=1}^K\gamma_k a_{x_k}^1\right)(\ell[\ell_1,0]+\ell[0,\ell_2]+\ell[\ell_1,\ell_2]), \nonumber
\end{align}
where $\ell, \ell_1, \ell_2 = 0, \ldots, L-1$. The left hand side is straightforward to estimate from data: it provides a succinct summary of it. The right hand side involves polynomial functions of unknowns $\gamma_1, \ldots, \gamma_K, x_1, \ldots, x_K$, and possibly $\sigma^2$. The task is to solve these polynomial equations in a robust way.

@@

\begin{proposition} \label{prop:poisson_model}
Under the Poisson process model, the first autocorrelation  is $\gamma a_y^1$, the second moment vector is $(\gamma a_y^1)^2 + \gamma a_y^2[\ell]$, and the third moment matrix is $(\gamma a_y^1)^3 + \gamma a_y^1  \cdot ( \gamma a_y^2[\ell_1] + \gamma a_y^2[\ell_2] + \gamma a_y^2[\ell_2-\ell_1]) + \gamma a_y^3(\ell_1,\ell_2) $. In particular, from the first three moments of the Poisson process model, once can recover the first three moments from the strongly-separated model, with the Poisson rate $\gamma$ playing the role of the ``occupancy factor''. So if recovery is possible for the strongly-separated model, it is also possible for the Poisson process model.
\end{proposition}

@@

Furthermore,
as proven in the following simple proposition.
%
\begin{proposition} \label{prop:uniqueness}
%
Let $z\in\RL$ and suppose that $z[0]$ and $z[L-1]$ are nonzero. Then $z$ is determined uniquely from  $a_z^2$ and $a_z^3$. More precisely, suppose we can measure $\tilde{a}_z^3[k,L-1] = a_z^3[k,L-1]+\upsilon$ and that $\vert z[0]z[L-1]\vert \geq \ell>0$.
Then,  $\hat{z}[k] =\frac{\tilde{a}_z^3[k,L-1]}{a_z^2[L-1]} $ satisfies $\vert \hat{z}[k] - z[k]\vert\leq \frac{\vert \upsilon\vert }{\ell}$.

@@

\begin{proof}
%
By assumption $a_z^2[L-1] = z[0]z[L-1]\neq 0$.
Then, the uniqueness results, for all $k=0,\ldots L-1$,  follows from:
%
\begin{equation*}
%
a_z^3[k,L-1] = z[0]z[k]z[L-1].
%
\end{equation*}
%
In addition, 
%
\begin{equation*}
%
\hat{z}[k] = \frac{\tilde{a}_z^3[k,L-1]}{a_z^2[L-1]} = z[k]+\frac{\upsilon}{a_z^2[L-1]} \quad \Rightarrow \quad \vert \hat{z}[k] - {z}[k]\vert \leq \frac{\vert\upsilon\vert}{\ell}.
%
\end{equation*} 
%
\end{proof}

@@

%\TODO{TB: I would place the equation counting argument in the theory section in the paragraph of open questions (I marked th place)}
%
Given the observation $y$, we proceed to compute the autocorrelations. The first-order autocorrelation is straightforward. For second-order autocorrelations, notice from equation~\eqref{eq:data_ac} that $a_y^2[\ell]$ suffers no noise-induced bias for $\ell$ in $1$ to $L-1$. Thus, we omit $\ell = 0$, which has the practical effect that we need not know $\sigma$ to make sense of the computed quantities. Likewise, for third-order autocorrelations, $a_y^3[\ell_1, \ell_2]$ for $0 \leq \ell_1, \ell_2 \leq L-1$ such that $\ell_2 \leq \ell_1$ includes all relevant entries for our purpose (this accounts for symmetries), and we further exclude any such that $\ell_1, \ell_2$ or $\ell_1 - \ell_2$ are zero to avoid the need to estimate $\sigma$---there are $\frac{(L-1)(L-2)}{2}$ remaining entries. We have
\begin{align*}
1 + (L-1) + \frac{(L-1)(L-2)}{2} = \frac{1}{2} L (L-1) + 1
\end{align*}
coefficients in total. Since we aim to estimate $KL$ parameters (for the $K$ signals of length $L$) plus $K$ parameters (for the densities $\gamma_k$), an absolute upper bound on $K$ (simply to ensure we have at least as many equations as we have unknowns) is
\begin{align*}
K(L+1) \leq \frac{1}{2} L (L-1) + 1.
\end{align*}
Thus, $(L-1)/2$ (up to a small approximation) is an absolute upper limit on $K$ (compare with~\cite{boumal2017heterogeneous,bandeira2017estimation}). 
In practice, the autocorrelations are computed on disjoint segments of $y$ of length $100\cdot10^6$ and added up, without correction for the junction points. Segments are handled sequentially on a GPU, as GPUs are particularly well suited to execute simple instructions across large vectors of data. If multiple GPUs are available, segments can of course be handled in parallel.

@@

@@

@@

We will also denote by $\L_l$ the autocorrelations of $X$:
%
\begin{align}
%
\LL_1 = \sum_{i=0}^{L-1} \M_1[i],
%
\end{align}
%
\begin{align}
%
\LL_2(\ell) = \sum_{i=0}^{L-1} \M_2[i,i+\ell],
%
\end{align}
%
and
%
\begin{align}
%
\LL_3(\ell_1,\ell_2) = \sum_{i=0}^{L-1} \M_3[i,i+\ell_1,i+\ell_2].
%
\end{align}

%
%Note that in the strongly-separated model, the first three observed moments are, respectively, $\LL_1$, $\LL_2(\ell)$, and $\LL_3(\ell_1,\ell_2)$.

@@

In this notation, we will show that the first moment of the data is $\gamma \LL_1$, the second moment vector is $(\gamma \LL_1)^2 + \gamma \LL_2(\ell)$, and the third moment matrix is $(\gamma \LL_1)^3 + \gamma \LL_1  \cdot ( \gamma\LL_2(\ell_1) + \gamma\LL_2(\ell_2) + \gamma\LL_2(\ell_2-\ell_1)) + \gamma \LL_3(\ell_1,\ell_2) $. In particular, from the first three moments of the Poisson process model, once can recover the first three moments from the well-separated model, with the Poisson rate $\gamma$ playing the role of the ``occupancy factor''. So if recovery is possible for the well-separated model, it is also possible for the Poisson process model.

@@

\paragraph{The first moment of $Y$.}

@@

\paragraph{The second moment of $Y$}

Again, we will condition on $M$ first, and then take the expectation over $M$. Fix $i_1 \ne i_2$, and let $\ell = i_2 - i_1$. Then:
%
\begin{align}
%
Y_{i_1} Y_{i_2} 
&= \sum_{j_1=0}^{L-1} \sum_{j_2=0}^{L-1} 
\sum_{k_1=1}^{M_{i_1-j_1}}\sum_{k_2=1}^{M_{i_2-j_2}}
X_{k_1}^{i_1-j_1}[j_1] X_{k_2}^{i_2 - j_2}[j_2].
%
\end{align}

We break up the double sum over $j_1$ and $j_2$ into two terms: one where $j_2 \ne j_1 + \ell$, and one where $j_2 = j_1 + \ell$ or equivalently $i_1-j_1 = i_2-j_2$. In the first case, all the terms are independent, and so the expectation factors. In the second case, when $k_1 \ne k_2$ we have independence, but otherwise not. This gives (all expectations are conditional on $M$):
